Paper

### 2024-12-17 paper/index.qmd
paper update

@@

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
source(file.path("..", "calibrate.R"))
# load(file.path("..", "out", "calibrate.RData"))
```

@@

$$ 
P_\text{de novo}(t) = 1 - \exp {( - r_0 T(t) )},
$$

@@

$$
P_\text{toilet}(t) = 1 - \exp ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
$$

@@

$$
\mathrm{T}(t) = \begin{cases}
    m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
    1 & \text{otherwise.}
\end{cases}
$$

@@

$$
X_i(t) = \max(X_\text{de novo}, X_\text{toilet}),
$$

@@

A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991;@keeling2011]:

$$
\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S- rS
$$

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected/colonised inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the proportion of new admissions that are susceptible/non-infected, while $(1 - p)$ is community colonisation/infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (_de novo_) colonisation/infection,
-   $D$ is the rate of patient discharges.

@@

Rejection sampling approximate Bayesian computation was employed to calibrate the model to the outbreak data [@sunnåker2013]. Priors were determined from domain knowledge or extracted from existing literature. The parameters in @tbl-consts were kept constant during the calibration process.

Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$. Next, each sample parameter point $\theta_i$ was used as an input to the agent-based model, generating a total of $n$ simulated datasets. The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$. Results from the first 30 simulated days were discarded to ensure that the model had reached a steady state.

@@

Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to generate a posterior distribution.

Firstly, the model was calibrated to pre-outbreak data to identify parameters that simulated the baseline rate. Next, we simulated the emergence of an outbreak and its subsequent control by incrementally raising colonisation pressure over time (via increased antimicrobial consumption, toilet use, etc.) and enhancing infection control measures, respectively. Finally, the posterior parameter distributions were used to simulate 20 outbreaks.

| Name                                                       | Value    | Description                                                             | Reference                |
|------------------|------------------|-------------------|------------------|
| Baseline (pre-outbreak) antimicrobial prescription rate    | 0.319    | Proportion of patients administered antimicrobials during admission     | @versporten2018          |
| Admission duration                                         | 8.3 days | Mean of a Poisson distribution                                          | @thehealthfoundation2023 |
| Proportion of toilets that are shared between \>1 patients | 0.6      | The remaining proportion (0.4) of toilets were dedicated to one patient | N/A                      |

: Parameters that were assumed to be constant during calibration. {#tbl-consts}

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures. Changes in parameters such as increased antimicrobial prescription rates and decreased toilet cleaning frequency were used to encourage increased disease transmission, leading to an outbreak. On the other hand, these parameters were reversed to control the outbreak.

## Model availability

@@

@@

```{r}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

single_seeds <- single_outbreak_sim$`random-seed`
single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
  filter(`random-seed` == single_seeds[1])

@@

# Initial conditions
initial_state <- c(N = 600, I = 17)  # Initial populations of non-infected and infected patients
time <- seq(0, run_to, by = 1)         # Time steps
dt <- 1                             # Time step size

@@

single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red")
```

>>>>

run_sims <- FALSE

@@

# to change body font size:
# change this line in typst-template.typ:
# set text(font: "STIX Two Text", size: 24pt)
library(tidyverse)
library(abc)
library(lhs)
library(flextable)
source(file.path("..", "calibrate.R"))
# load(file.path("..", "out", "calibrate.RData"))
```

@@

$$ 
P_\text{de novo}(t) = 1 - \text{Exp} {( - r_0 T(t) )},
$$

@@

$$
P_\text{toilet}(t) = 1 - \text{Exp} ( {-\sum_{j=0}^{U_i(t)}[C(t_j) \cdot T(t)]} )
$$

@@

$$
T(t) = \begin{cases}
    m & \text{if on antimicrobial therapy} \\ % & is your "\tab"-like command (it's a tab alignment character)
    1 & \text{otherwise.}
\end{cases}
$$

@@

$$
X_i(t) = \text{Max}(X_\text{de novo}, X_\text{toilet}),
$$

@@

A deterministic representation of the model was described in terms of the following differential equations, an adaptation of *SI* (susceptible--infection) models [@anderson1991; @keeling2011]:

$$
\frac{dS}{dt} = pA - D \cdot \frac{S}{S + I} - \beta S - rS
$$

@@

-   $dS/dt$ is the rate of change of susceptible/non-infected inpatients,
-   $S$ is the number of susceptible inpatients,
-   $I$ is the number of infected/colonised inpatients,
-   $A$ is the rate of new admissions,
-   $p$ is the proportion of new admissions that are susceptible/non-infected, while $(1 - p)$ is community colonisation/infection rate,
-   $\beta$ is the transmission coefficient,
-   $r$ is the rate of random (*de novo*) colonisation/infection,
-   $D$ is the rate of patient discharges.

@@

Rejection sampling approximate Bayesian computation (ABC) was employed to calibrate the model to the outbreak data [@sunnåker2013]. Priors were determined from domain knowledge or extracted from existing literature. A summary of the priors used is available in @tbl-priors.

Firstly, latin hypercube sampling was used to generate $n$ samples from the parameter prior distributions, using a uniform distribution, such that $\theta_1,\theta_2,\ldots,\theta_n$. Next, each sample parameter point $\theta_i$ was used as an input to the agent-based simulation model, generating a total of $n$ simulated datasets. The number of daily cases of nosocomial CDI were extracted from the model outputs, and converted to a time series of nosocomial CDI per 1000 patient days for each month, generating simulated summary statistics $S_1,S_2,\ldots,S_n$. Results from the first 31 simulated days were discarded to ensure that the model had reached a steady state.

@@

Otherwise, $S_i$ was rejected. The parameter samples that generated the accepted simulations were used to produce a posterior distribution. $\epsilon$ was set so that the acceptance rate was approximately 2.5%.

The approach to calibration was sequential. Firstly, ABC was used to calibrate the model pre-outbreak data to identify parameters that simulated the baseline rate. Next, the posterior distribution of the baseline rate model was used as a prior for a model that extended the simulation into the period of CDI outbreak and implementation of enhanced IPC measures (October 2004--May 2005 in the observed dataset) [@salgado2009]. To simulate the emergence of an outbreak, the model changed parameters at a discrete time point corresponding to the outbreak onset. The priors used for ABC were selected to be compatible with increased transmission or colonisation rates, for example increased antimicrobial exposure and decreased toilet cleaning frequency. At a subsequent discrete time point, the model again changed parameters to simulate enhanced infection control measures. The prior distributions for these parameters were selected to decrease transmission and colonisation -- generally the reverse of the 'outbreak' parameters.

The posterior parameter distributions were used to simulate 20 outbreak scenarios. Parameter samples were drawn from a uniform distribution:

$$
U_[a,b] = U(\mu - \sigma, \mu + \sigma),
$$

where: $\mu$ and \$\sigma\$ are the mean and standard deviation of the posterior distribution, respectively. The simulated rates of nosocomial CDI were compared to the observed rates to assess the model's performance. One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method. Since the agent-based model was stochastic, all simulations were aggregated

| Name                                                       | Distribution | Parameters                                                                                                   | Description                                                                     | Reference                                          |
|---------------|---------------|---------------|---------------|---------------|
| Baseline (pre-outbreak) antimicrobial prescription rate    | Constant     | 49.9%                                                                                                        | Proportion of patients administered antimicrobials during admission             | @versporten2018                                    |
| Admission duration                                         | Poisson      | $\lambda = 4.6  \text{ days}$                                                                                |                                                                                 | @levitkryankelixhauserastrangesekassedccoffeyr2007 |
| Proportion of toilets that are shared between \>1 patients | Constant     | 60%                                                                                                          | The remaining proportion (40%) of toilets were dedicated to one patient         | N/A                                                |
| Community colonisation rate                                | Uniform      | $a = 0.01, b = 0.1$                                                                                          |                                                                                 | @crobach2018                                       |
| Antimicrobial effect                                       | Uniform      | $ a = 1.31, b = 1.87 $                                                                                     | Odds ratio for effect of antimicrobial therapy on risk of hospital-acquired CDI | @slimings2014                                      |
| CDI incidence                                              | Uniform      | $a=2.8, b=12.1$                               | Per 10,000 patient-days                                                         | @marra2020                                         |
| Mean toilet frequenting rate                               | Uniform      | $a=0.5,b=5$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
| Mean toilet cleaning rate                                  | Uniform      | $a=0.5, b=3$                               | Per day; used as \lambda parameter for Poisson distribution                     |                                                    |
| Toilet contamination effect                                | Uniform      | $a=0.01,b=0.99$                               | Contamination impact by infected/colonised patients through toilet use          |                                                    |
| Toilet cleaning effect                                     | Uniform      | $a=0.01, b=0.99$                               | Impact of toilet cleaning on contamination                                      |                                                    |

: Parameters that were assumed to be constant during calibration. {#tbl-priors}

@@

To simulate the generation of outbreaks and their subsequent control through enhanced infection prevention and control (IPC) measures, the agent-based model optionally changed its parameters at four configurable discrete time points within the simulation: 1) onset of outbreak; 2) start of enhanced IPC measures; 3) end of outbreak; and 4) end of enhanced IPC measures. During calibration, the upper or lower bounds for parameter priors during the outbreak period were extended by 50% of baseline. For example, the upper bound of prior for the parameter determining the rate of antimicrobial prescriptions was 50% higher than the baseline rate. On the other hand, these parameters were reversed to control the outbreak.

## Software implementation

The agent-based model was implemented in NetLogo v6.2.2 [@wilenskyu.1999]. The R package `nlrx` was used to execute simulations and transfer output data for analysis in R [@salecker2019; @rcoreteam2022].

@@

```{r, include=T}
params_table <- abc_params_outbreak_control$unadj.values %>% 
  as_tibble() %>% 
  dplyr::select(!any_of(c("outbreak-start", "outbreak-end", "control-end"))) %>%
  dplyr::rename_with(.fn = \(.x) paste0("b-", .x), .cols = !starts_with("o-") & !starts_with("c-")) %>% 
  summarise(across(everything(), ~ paste0(
    round(mean(.x), 2),
    " (",
    round(sd(.x), 2),
    ")"
    )))

param_units <- c("toilet-frequenting-rate" = "times per day",
                 "toilet-cleaning-rate" = "times per day",
                 "toilet-contamination-effect" = "",
                 "toilet-cleaning-effect" = "",
                 "community-colonisation-rate" = "proportion of patients",
                 "antibiotic-prescription-rate" = "proportion of admissions",
                 "antibiotic-effect" = "relative risk",
                 "random-colonisation" = "cases per 10,000 bed days")

param_names <- c("toilet-frequenting-rate" = "Toilet frequenting rate",
                 "toilet-cleaning-rate" = "Toilet cleaning rate",
                 "toilet-contamination-effect" = "Toilet contamination effect",
                 "toilet-cleaning-effect" = "Toilet cleaning effect",
                 "community-colonisation-rate" = "Community colonisation rate",
                 "antibiotic-prescription-rate" = "Antibiotic prescription rate",
                 "antibiotic-effect" = "Antibiotic effect",
                 "random-colonisation" = "Random colonisation")

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  flextable() %>% 
  labelizor(j = "units", part = "body",
            labels = param_units) %>% 
  labelizor(j = "param", part = "body",
            labels = param_names)

params_table %>% 
  pivot_longer(cols = everything(), names_to = c("period", "param"), values_to = "value",
               names_pattern = "(b|o|c)-(.*)") %>% 
  pivot_wider(names_from = period, values_from = value) %>% 
  # reorder cols
  dplyr::select(param, b, o, c) %>% 
  dplyr::rename(baseline = b, outbreak = o, control = c) %>% 
  mutate(units = param) %>% 
  relocate(units, .after = param) %>%
  knitr::kable()
```

@@

```{r, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(single_outbreak_sim$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

single_seeds <- unique(single_outbreak_sim$`random-seed`)
single_outbreak_one_seed <- single_outbreak_sim_mod %>% 
  filter(`random-seed` == single_seeds[3])

@@

# Initial conditions
total_patients <- unique(single_outbreak_one_seed$`current-inpatients`)
init_N <- unique(single_outbreak_one_seed$`community-colonisation-rate`) * total_patients
init_I <- total_patients - init_N
initial_state <- c(N = init_N, I = init_I)  # Initial populations of non-infected and infected patients
time <- seq(0, run_to, by = 1)         # Time steps
dt <- 1                             # Time step size

@@

single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red") +
  xlim(500, 2000)
```
